---
title: "Sample Size Planning with sspLNIRT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample Size Planning with sspLNIRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `sspLNIRT` package provides sample size planning tools for item calibration
under the Joint Hierarchical Model (JHM; van der Linden, 2007). The JHM
combines a two-parameter normal ogive model for response accuracy with a
log-normal model for response times, estimated via Gibbs sampling through the
`LNIRT` package (Fox et al., 2023).

The central question the package addresses is: *Given assumptions about the
data-generating process, what is the minimum number of respondents needed to
achieve a desired level of item parameter accuracy?* Accuracy is defined as the
root mean squared error (RMSE) of a target item parameter falling below a
user-specified threshold.

The package supports three workflows, each described in a separate section of
this vignette:

1. **Shiny app** — an interactive interface for exploring precomputed results
   or generating a customized R script.
2. **R package with precomputed results** — programmatic access to a database
   of 2,400 precomputed design conditions, with functions for retrieval,
   inspection, and visualization.
3. **Custom simulations** — running `comp_rmse()` or `optim_sample()` for
   design conditions not covered by the precomputed grid.

```{r install, eval = FALSE}
# Install from GitHub
if (!requireNamespace("devtools")) install.packages("devtools")
devtools::install_github("anonymous-peer-2026/sspLNIRT")
```

```{r setup}
library(sspLNIRT)
```


## 1. Using the Shiny App

The package includes an interactive Shiny application that provides the most
accessible entry point. Launch it with:

```{r, eval = FALSE}
run_app()
```

The app allows users to:

- Select design parameters (test length, mean item discrimination, residual
  variance level, ability--speed correlation) and target RMSE thresholds via
  dropdown menus.
- Retrieve the corresponding minimum sample size from the precomputed database,
  together with estimation metrics (RMSE, Monte Carlo SD, bias) for all item
  and person parameters.
- Inspect diagnostic plots including estimation accuracy across parameter
  values, power curves, convergence diagnostics, and simulated response
  accuracy and response time distributions.
- Download the result objects (`.rds`) for further use in R.

For design conditions not covered by the precomputed grid, the app can generate
a customized R script with the appropriate function calls, which can then be
executed locally or on a computing cluster. Detailed documentation of the app
interface is available at <https://github.com/anonymous-peer-2026/sspLNIRT/>.


## 2. Using Precomputed Results in R

This section walks through the full workflow for retrieving and inspecting
precomputed results programmatically. The precomputed database covers 2,400
design conditions; if your specification falls within this grid, no simulations
are needed.

### 2.1 Check available configurations

Start by inspecting which design conditions are available in the precomputed
grid using `available_configs()`:

```{r}
configs <- available_configs()
head(configs, 10)
```

Each row represents one precomputed optimization run. The columns correspond to
the design factors that were systematically varied: `thresh` (RMSE threshold),
`out.par` (target item parameter), `K` (test length), `mu.alpha` (mean item
discrimination), `meanlog.sigma2` (log-scale mean of the residual variance
$\sigma^2$), and `rho` (ability--speed correlation). All other model parameters
(e.g., mean item difficulty, item parameter correlations, standard deviations)
were held constant across the grid. Their values are stored in the `design`
element of each result (see Section 2.3).

To check the available levels per factor:

```{r}
unique(configs$out.par)
unique(configs$thresh)
unique(configs$K)
unique(configs$mu.alpha)
unique(configs$meanlog.sigma2)
unique(configs$rho)
```

### 2.2 Retrieve the minimum sample size

The `get_sspLNIRT()` function looks up precomputed results by matching on the
six design factors. It accepts vectors for `thresh` and `out.par` to target
multiple item parameters simultaneously:

```{r}
result <- get_sspLNIRT(
  thresh         = c(0.10, 0.20, 0.05, 0.05),
  out.par        = c("alpha", "beta", "phi", "lambda"),
  K              = 30,
  mu.alpha       = 1,
  meanlog.sigma2 = log(0.6),
  rho            = 0.4
)
```

Because the precomputed database stores results for each target parameter
separately, `get_sspLNIRT()` matches each `out.par`/`thresh` pair
independently. It then returns the bottleneck result — the parameter that
required the largest sample size, since that $N$ guarantees all other
parameters also meet their (less demanding) thresholds.

The `summary()` method provides a concise overview:

```{r}
summary(result$object)
```

For planning based on a single parameter, scalar inputs work the same way:

```{r}
res_alpha <- get_sspLNIRT(
  thresh         = 0.10,
  out.par        = "alpha",
  K              = 30,
  mu.alpha       = 1,
  meanlog.sigma2 = log(0.6),
  rho            = 0.4
)
summary(res_alpha$object)
```

### 2.3 Understand the output

The object returned by `get_sspLNIRT()` is a list with two elements:

- **`result$object`**: the optimization result (class `sspLNIRT.object`),
  containing:
  - `N.min` — the minimum sample size identified by the bisection optimizer.
    If the lower bound already satisfied the threshold, this is the string
    `"res.lb < thresh"`; if the upper bound was insufficient, `"res.ub >
    thresh"`.
  - `res.best` — the RMSE of the target parameter at `N.min`.
  - `comp.rmse` — the full accuracy evaluation at `N.min`, including RMSE, MC
    SD, and bias for all item and person parameters, as well as binned error
    data and convergence diagnostics.
  - `trace` — the optimization trace: sample sizes and RMSE values at each
    bisection step, the number of steps, and the elapsed time.

- **`result$design`**: the full set of input parameters used for the
  precomputation (class `sspLNIRT.design`). This includes the constant
  parameters that were not varied across the grid:

```{r}
str(result$design, max.level = 1)
```

Notably, `result$design$out.par` and `result$design$thresh` identify which
single-parameter configuration was the bottleneck:

```{r}
result$design$out.par
result$design$thresh
```

This tells you which parameter drove the minimum sample size. In the
multi-parameter call above, the returned `N.min` and all diagnostics correspond
to the optimization run for this bottleneck parameter.

### 2.4 Inspect the implied distributions

With the design object in hand, you can inspect what the assumed parameter
values imply for observable data. The `plot_RA()` and `plot_RT()` functions
simulate data under the specified model and display the resulting
distributions. Using `result$design` avoids having to re-specify all the
constant parameters manually:

```{r}
cfg <- result$design
sim.mod.args <- list(
  K              = cfg$K,
  mu.person      = cfg$mu.person,
  mu.item        = cfg$mu.item,
  meanlog.sigma2 = cfg$meanlog.sigma2,
  sdlog.sigma2   = cfg$sdlog.sigma2,
  cov.m.person   = cfg$cov.m.person,
  cov.m.item     = cfg$cov.m.item,
  sd.item        = cfg$sd.item,
  cor2cov.item   = cfg$cor2cov.item
)
```

**Item-level response accuracy.** With `by.theta = TRUE`, `plot_RA()` shows
item characteristic curves (ICCs) — the probability of a correct response as
a function of ability $\theta$. Each panel corresponds to one item. The dashed
vertical line marks the item difficulty ($\beta$), and the faint dots show
simulated binary responses:

```{r, fig.width = 7, fig.height = 5}
do.call(plot_RA, c(list(level = "item", by.theta = TRUE, N = 1e4), sim.mod.args))
```

Setting `by.theta = FALSE` shows the marginal distribution of response
probabilities across persons for each item:

```{r, fig.width = 7, fig.height = 5}
do.call(plot_RA, c(list(level = "item", by.theta = FALSE, N = 1e4), sim.mod.args))
```

**Person-level response accuracy.** The total correct score distribution
provides a summary of expected test performance:

```{r, fig.width = 7, fig.height = 4}
do.call(plot_RA, c(list(level = "person", by.theta = FALSE, N = 1e4), sim.mod.args))
```

The `by.theta = TRUE` variant shows mean total score as a function of ability:

```{r, fig.width = 7, fig.height = 4}
do.call(plot_RA, c(list(level = "person", by.theta = TRUE, N = 1e4), sim.mod.args))
```

**Item-level response times.** Each panel shows the RT density for one item.
The `logRT` argument toggles between seconds and the log scale:

```{r, fig.width = 7, fig.height = 5}
do.call(plot_RT, c(list(level = "item", logRT = FALSE, N = 1e4), sim.mod.args))
```

```{r, fig.width = 7, fig.height = 5}
do.call(plot_RT, c(list(level = "item", logRT = TRUE, N = 1e4), sim.mod.args))
```

**Person-level response times.** Quantile reference lines summarize the
distribution of mean response times across persons:

```{r, fig.width = 7, fig.height = 4}
do.call(plot_RT, c(list(level = "person", logRT = TRUE, N = 1e4), sim.mod.args))
```

```{r, fig.width = 7, fig.height = 4}
do.call(plot_RT, c(list(level = "person", logRT = FALSE, N = 1e4), sim.mod.args))
```

These plots help evaluate whether the assumed data-generating process produces
plausible score and time distributions for the intended application. They may
also inform adjustments to the input parameters before committing to a sample
size.

### 2.5 Inspect estimation accuracy

The `plot_estimation()` function displays RMSE or bias as a function of the
true (simulated) parameter values at the minimum $N$. This can quantify to what 
extent estimation accuracy varies systematically across the parameter range. For
instance, how much items with extreme difficulty values tend to be estimated 
less precisely.

For item parameters:

```{r, fig.width = 7, fig.height = 5}
plot_estimation(result$object, pars = "item", y.val = "rmse")
```

```{r, fig.width = 7, fig.height = 5}
plot_estimation(result$object, pars = "item", y.val = "bias")
```

For person parameters:

```{r, fig.width = 7, fig.height = 4}
plot_estimation(result$object, pars = "person", y.val = "rmse")
```

Each point represents the mean value within a quantile bin of the true
parameter. The smooth line is a LOESS fit.

### 2.6 Inspect the power curve

The `plot_power_curve()` function fits a log-log regression through the
bisection trace and displays the relationship between sample size and RMSE:

```{r, fig.width = 7, fig.height = 4}
plot_power_curve(result$object, out.par = result$design$out.par,
                 thresh = result$design$thresh)
```

The dashed horizontal line marks the RMSE threshold; the dotted vertical line
marks the minimum $N$. The left panel shows the log-log fit; the right panel
shows the same relationship on the original scale. The $R^2$ of the log-log
regression is reported in the caption.

### 2.7 Inspect convergence diagnostics

Convergence of the MCMC chains can be assessed from the $\hat{R}$ diagnostics
stored in the result object. These are quantiles of the $\hat{R}$ statistic
(Vehtari et al., 2021) across Monte Carlo iterations:

```{r}
rhat <- result$object$comp.rmse$rhat.dat
if (!is.null(rhat)) {
  if (is.data.frame(rhat) || is.matrix(rhat)) {
    round(rhat, 4)
  }
}
```

Values of 1 indicate that the chains have mixed well. The rows represent
quantiles (50th, 80th, 90th, 95th percentile); the columns correspond to item
parameters.


## 3. Custom Simulations

When the intended design is not covered by the precomputed grid, the package
provides two functions for running simulations directly:

- `comp_rmse()` evaluates item parameter accuracy at a single, fixed sample
  size.
- `optim_sample()` runs the bisection optimizer to find the minimum sample size
  for a target RMSE threshold.

Both functions are computationally intensive (minutes to hours depending on
settings) and benefit from parallel execution.

### 3.1 Setup

The simulation and estimation are parallelized over Monte Carlo replications
using the `future` framework. Set up a parallel backend before calling either
function:

```{r, eval = FALSE}
library(future)
plan(multisession, workers = 4)
```

### 3.2 Evaluate accuracy at a fixed N

`comp_rmse()` simulates data, fits the model, and computes RMSE, MC SD, and
bias for all item and person parameters at a specified sample size:

```{r, eval = FALSE}
rmse_result <- comp_rmse(
  N              = 200,
  iter           = 200,
  K              = 30,
  mu.person      = c(0, 0),
  mu.item        = c(1, 0, 0.5, 1),
  meanlog.sigma2 = log(0.6),
  cov.m.person   = matrix(c(1, 0.4, 0.4, 1), ncol = 2),
  cov.m.item     = matrix(c(1, 0, 0, 0,
                             0, 1, 0, 0.4,
                             0, 0, 1, 0,
                             0, 0.4, 0, 1), ncol = 4),
  sd.item        = c(0.2, 1, 0.2, 0.5),
  cor2cov.item   = TRUE,
  sdlog.sigma2   = 0,
  XG             = 5000,
  burnin         = 20,
  seed           = 1234,
  keep.err.dat   = TRUE,
  keep.rhat.dat  = TRUE
)

summary(rmse_result)
```

Key arguments:

- `iter`: number of Monte Carlo replications (simulated datasets). Higher
  values reduce the MC standard deviation of the RMSE estimate.
- `XG`: number of Gibbs sampler iterations per chain (4 chains are run per
  replication).
- `keep.err.dat`: if `TRUE`, retains the full per-replication error data
  (needed for `plot_estimation()` with custom binning). If `FALSE`, errors are
  pre-binned into `K` quantile bins.
- `keep.rhat.dat`: if `TRUE`, retains the $\hat{R}$ matrix across
  replications.
- `seed`: passed to `future.apply` for parallel-safe reproducibility.

The output has the same `sspLNIRT.object` class as `optim_sample()` results and
works with `summary()`, `plot_estimation()`, and `print()`.

### 3.3 Optimize for the minimum sample size

`optim_sample()` wraps `comp_rmse()` in a bisection loop. It accepts vectors
for `thresh` and `out.par` to simultaneously target multiple item parameters:

```{r, eval = FALSE}
custom_result <- optim_sample(
  thresh         = c(0.10, 0.15),
  out.par        = c("alpha", "beta"),
  range          = c(50, 2000),
  iter           = 200,
  K              = 30,
  mu.person      = c(0, 0),
  mu.item        = c(1, 0, 0.5, 1),
  meanlog.sigma2 = log(0.6),
  cov.m.person   = matrix(c(1, 0.4, 0.4, 1), ncol = 2),
  cov.m.item     = matrix(c(1, 0, 0, 0,
                             0, 1, 0, 0.4,
                             0, 0, 1, 0,
                             0, 0.4, 0, 1), ncol = 4),
  sd.item        = c(0.2, 1, 0.2, 0.5),
  cor2cov.item   = TRUE,
  sdlog.sigma2   = 0,
  XG             = 5000,
  burnin         = 20,
  seed           = 1234,
  keep.err.dat   = FALSE,
  keep.rhat.dat  = TRUE
)

summary(custom_result)
```

The `range` argument sets the search interval for $N$. The bisection halves the
interval at each step, evaluating `comp_rmse()` at the midpoint, until no
further refinement is possible (i.e., the interval width is one). The number of
bisection steps is approximately $\lceil \log_2(\text{range}[2] -
\text{range}[1]) \rceil$, each requiring a full `comp_rmse()` evaluation.

When multiple parameters are targeted, the bisection requires *all* specified
parameters to simultaneously fall below their respective thresholds before
narrowing the search interval, so the resulting $N$ satisfies all targets at
once. This differs from the `get_sspLNIRT()` approach, which looks up each
parameter independently and selects the bottleneck post hoc.

The result object supports the same inspection functions demonstrated in
Section 2:

```{r, eval = FALSE}
plot_estimation(custom_result, pars = "item", y.val = "rmse")
plot_power_curve(custom_result, out.par = "alpha", thresh = 0.10)
```

### 3.4 Practical considerations

- **Computation time**: a single `comp_rmse()` call with `iter = 200` and
  `XG = 5000` can take minutes to hours depending on $N$, $K$, and available
  cores. The full bisection typically requires 10--13 evaluations, so
  `optim_sample()` can take several hours.
- **Parallelization**: setting `future::plan(multisession)` with multiple
  workers distributes replications across cores. The number of workers should
  match available cores.
- **Reproducibility**: the `seed` argument ensures parallel-safe
  reproducibility via L'Ecuyer-CMRG streams. Pass an explicit integer for full
  reproducibility.
- **Memory**: with `keep.err.dat = TRUE`, the full error arrays are retained in
  memory. For large $N$ and $K$, setting this to `FALSE` reduces memory usage
  by pre-binning errors.

```{r, eval = FALSE}
# Reset to sequential when done
plan(sequential)
```


## References

Fox, J.-P., Klotzke, K., & Simsek, A. S. (2023). R-package LNIRT for joint
modeling of response accuracy and times. *PeerJ Computer Science*, *9*, e1232.
<https://doi.org/10.7717/peerj-cs.1232>

van der Linden, W. J. (2007). A hierarchical framework for modeling speed and
accuracy on test items. *Psychometrika*, *72*(3), 287--308.
<https://doi.org/10.1007/s11336-006-1478-z>

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P.-C. (2021).
Rank-normalization, folding, and localization: An improved $\hat{R}$ for
assessing convergence of MCMC (with discussion). *Bayesian Analysis*, *16*(2),
667--718.
<https://doi.org/10.1214/20-BA1221>
